Pre working group data exploration: Mainly wrt coral cover

### 2024-07-09 julianna_data_exploration.Rmd
Did more analyses; worked towards Deron's vision for "figure 1"

# load packages
require(here) # relative file paths
require(tidyverse) # data wrangling
require(vegan) # community/diversity analyses
require(lme4) # glmer()
require(emmeans) # emmeans()
```

@@

```{r}
allbenthos <- read_csv(here("data", "allbenthicdata_cleaned_colnames.csv")) 
algae <- read_csv(here("data", "algaeonly_cleaned_colnames.csv"))
```

@@

```{r}
# matrix of just species and their relative percent covers in the same order as the algae tibble
algae %>% 
  select(-year, -location, -site, -habitat, -site_habitat, -transect, -quadrat) -> sppMat
```

```{r}
algae %>% 
  # grab the metadata info that is in the same order (can re-join after)
  select(year, location, site, habitat, site_habitat, transect, quadrat) %>% 
  mutate(Richness = specnumber(sppMat),
         Shannon = diversity(sppMat, "shannon"),
         Richness = specnumber(sppMat)) -> algDiv
```

@@

@@

@@

```{r}
allbenthos %>% 
  select(year, location, site, habitat, site_habitat, transect, quadrat, coral) %>% 
  full_join(algDiv) %>% 
  # add column for each transect ID
  rowwise() %>% 
  mutate(transect_id = paste0(site, habitat, year, transect)) -> coralAlgDiv  
```

## Plot shannon

Plot Shannon diversity across sites and habitats--what's up with this structure? Emailed Jamie and asked Lauren. TBD

```{r}
coralAlgDiv %>% 
  ggplot(aes(x = coral, y = Shannon, color = year)) +
  geom_point() + 
  facet_grid(habitat~site) +
  theme_classic()
```

@@

@@

```{r}
coralAlgDiv %>% 
  ggplot(aes(x = Richness, y = coral, color = year, group = Richness)) +
  geom_boxplot() + 
  facet_grid(site ~ habitat) +
  theme_classic()
```

@@

```{r}
m1 <- glmer(Richness ~ scale(coral) + habitat + (1|site) + (1|transect_id) + (1|year), 
            family = poisson(link = "log"), data = coralAlgDiv)
  plot(m1)
  summary(m1)
  emmeans(m1, ~ habitat)
```

@@

```{r}
# spp matrix
algae %>% 
  mutate(AlgSum = rowSums(across(where(is.numeric)))) %>% 
  filter(AlgSum != 0) %>% 
  select(-year, -location, -site, -habitat, -site_habitat, -transect, -quadrat, -AlgSum) -> sppMat0

# metadata matrix
algae %>% 
  mutate(AlgSum = rowSums(across(where(is.numeric)))) %>% 
  filter(AlgSum != 0) %>%
  select(year, location, site, habitat, site_habitat, transect, quadrat) -> sppMeta0

@@

```{r}
siteHabBetaDisper <- vegan::betadisper(bray_matrix, sppMeta0$site_habitat)

# plot of dispersion test
plot(year_beta_disper, sub = "") # lol
boxplot(year_beta_disper, main = "Year", xlab = "") 
```

>>>>

# load packages
require(here) # relative file paths
require(tidyverse) # data wrangling
require(vegan) # community/diversity analyses
require(lme4) # glmer()
require(emmeans) # emmeans()
require(viridis) # color palette
require(MoMAColors) # another color palette
```

@@

```{r}
allbenthos <- read_csv(here("data", "allbenthicdata_cleaned_colnames.csv")) 
algae <- read_csv(here("data", "algaeonly_cleaned_colnames.csv")) %>% 
  mutate(island_side = case_when(site == "lter_1" ~ "north",
                                 site == "lter_2" ~ "north",
                                 site == "lter_3" ~ "east",
                                 site == "lter_4" ~ "east",
                                 site == "lter_5" ~ "west",
                                 site == "lter_6" ~ "west",
                                 TRUE ~ "UdINGLEbUTT"))
  # going to run everything without algal turf
  # select(-algal_turf)

macroalgae <- algae %>% 
  select(-algal_turf, -crustose_corallines, -cyanophyta, -damselfish_turf, -lithophyllum_kotschyanum,
         -neogoniolithon_brassica_florida, -sporolithon_sp) # got this from the cleaning code--should Peysonellia be included?
```

@@

```{r}
# matrix of just species and their relative percent covers in the same order as the algae tibble
algae %>% 
  select(-year, -location, -site, -island_side, -habitat, -site_habitat, -transect, -quadrat) -> sppMat
```

```{r}
algae %>% 
  # grab the metadata info that is in the same order (can re-join after)
  select(year, location, site, island_side, habitat, site_habitat, transect, quadrat) %>% 
  mutate(Richness = specnumber(sppMat),
         Shannon = diversity(sppMat, "shannon"),
         Evenness = Shannon/log(Richness), # I'm pretty sure this is right but double check (Pielou's evenness)
         Richness = specnumber(sppMat)) %>% 
  # clean for graphing
  mutate(habitat = factor(habitat, levels = c("fringing", "backreef", "outer_10", "outer_17")),
         island_side = factor(island_side, levels = c("north", "east", "west"))) -> algDiv
```

@@

Island side:

```{r}
algDiv %>% 
  ggplot(aes(x = island_side, y = Shannon, group = island_side, fill = island_side)) +
  geom_boxplot() +
  facet_wrap(~habitat) +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust=1))

@@

Broken up by island side

```{r}
algDiv %>% 
  mutate(habitat = factor(habitat, levels = c("fringing", "backreef", "outer_10", "outer_17"))) %>% 
  ggplot(aes(x = Richness, group = habitat, fill = habitat)) +
  geom_bar(position = "dodge") +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust=1)) +
  scale_fill_viridis_d() +
  facet_wrap(~island_side)
```

# Compare algal and coral cover

```{r}
# get ranges to calculate over
startAlg <- grep("acanthophora_spicifera", colnames(algae))
endAlg <- grep("valonia_ventricosa", colnames(algae))

algae %>% 
  mutate(AlgSum = rowSums(algae[startAlg:endAlg])) -> algSum

algSum %>% 
  full_join(allbenthos) %>% 
  ggplot(aes(x = coral, y = AlgSum, color = year)) +
  geom_point() +
  facet_grid(site~habitat)

@@

```{r}
allbenthos %>% 
  select(year, location, site, habitat, site_habitat, transect, quadrat, coral) %>% 
  full_join(algDiv) %>% 
  # add column for each transect ID
  rowwise() %>% 
  mutate(transect_id = paste0(site, habitat, "_", year, "_", transect)) %>% 
  # clean for graphing
  mutate(habitat = factor(habitat, levels = c("fringing", "backreef", "outer_10", "outer_17")),
         island_side = factor(island_side, levels = c("north", "east", "west"))) -> coralAlgDiv  
```

## Plot Shannon

Plot Shannon diversity across sites and habitats--what's up with this structure?

```{r}
coralAlgDiv %>% 
  ggplot(aes(x = coral, y = Shannon, color = as.factor(Richness))) +
  geom_point(alpha = 0.5) + 
  facet_grid(habitat~site) +
  theme_classic() + 
  geom_hline(yintercept = 0.693, color = "red", linetype = "dashed") # this is when two species are equally distributed (half half)--not sure why that would be
```

@@

Or look at Shannon vs. the percent cover of algae in the plot

```{r}
coralAlgDiv %>% 
  full_join(algSum) %>% 
  ggplot(aes(x = AlgSum, y = Shannon, color = as.factor(Richness))) +
  geom_point(alpha = 0.5) + 
  facet_grid(habitat~site) +
  theme_classic() + 
  geom_hline(yintercept = 0.693, color = "red", linetype = "dashed")
```

What about this when Richness == 2?

```{r}
coralAlgDiv %>% 
  full_join(algSum) %>% 
  filter(Richness == 2) %>% 
  ggplot(aes(x = AlgSum, y = Shannon, color = as.factor(Richness))) +
  geom_point(alpha = 0.5) + 
  facet_grid(habitat~site) +
  theme_classic() + 
  geom_hline(yintercept = 0.693, color = "red", linetype = "dashed")
```

Could this be because the only way to get a high enough imbalance/unevenness is having a large enough algal N

What does it look like relative to turf algae?

```{r}
coralAlgDiv %>% 
  full_join(algSum) %>% 
  filter(Richness == 2) %>% 
  ggplot(aes(x = algal_turf, y = Shannon, color = as.factor(Richness))) +
  geom_point(alpha = 0.5) + 
  facet_grid(habitat~site) +
  theme_classic() + 
  geom_hline(yintercept = 0.693, color = "red", linetype = "dashed")
```

Nic: are these lines all turf but 1 line, all turf but 2 points line, all turf but 3 points line?

```{r}
coralAlgDiv %>% 
  full_join(algSum) %>% 
  filter(Richness == 2) %>% 
  mutate(Nonturf = AlgSum - algal_turf) %>% 
  ggplot(aes(x = AlgSum, y = Shannon, color = as.factor(Nonturf))) +
  geom_point(alpha = 0.5) + 
  facet_grid(habitat~site) +
  theme_classic() + 
  geom_hline(yintercept = 0.693, color = "red", linetype = "dashed")
```

^This seems to be true here. It seems like lines have to do with how much of the non-turf you have. They only do 25 points and multiply by 4 so everything is a multiple of 4. But then how can we have a 2 there??? Answer: 2020 averaging

@@

```{r}
coralAlgDiv %>% 
  ggplot(aes(x = Richness, y = coral, fill = habitat, group = Richness)) +
  geom_boxplot() + 
  facet_grid(island_side ~ habitat) +
  scale_fill_viridis_d() +
  theme_classic()
```

@@

```{r}
m1 <- glmer(Richness ~ scale(coral) + habitat + island_side + (1|site) + (1|transect_id) + (1|year), 
            family = poisson(link = "log"), data = coralAlgDiv)
  plot(m1) # I think this looks pretty weird right?
  summary(m1)
  emmeans(m1, ~ habitat)
```

## Plot evenness

Plot evennes across habitats and years

```{r}
coralAlgDiv %>% 
  ggplot(aes(x = coral, y = Evenness, color = year, group = Evenness)) +
  geom_jitter() + 
  facet_grid(habitat~site) +
  theme_classic()
```

# Plot changes of algae, coral, etc. through time

```{r make a big joined tibble}
startMAlg <- grep("acanthophora_spicifera", colnames(macroalgae))
endMAlg <- grep("valonia_ventricosa", colnames(macroalgae))

macroalgae %>% 
  mutate(macroalgae_sum = rowSums(macroalgae[startMAlg:endMAlg])) %>% 
  select(year, location, site, habitat, site_habitat, transect, quadrat, island_side, macroalgae_sum) %>% 
  full_join(coralAlgDiv) %>% 
  full_join(algae) %>% 
  rowwise() %>% 
  mutate(turf_sum = sum(algal_turf, damselfish_turf),
         CCA_sum = sum(lithophyllum_kotschyanum, crustose_corallines, neogoniolithon_brassica_florida, sporolithon_sp)) %>% 
  select(year, location, site, habitat, site_habitat, transect, quadrat, island_side, turf_sum, CCA_sum, coral, macroalgae_sum) %>% 
  # pivot key columns longer
  pivot_longer(cols = c(turf_sum, CCA_sum, coral, macroalgae_sum),
               names_to = "substrate",
               values_to = "sum_cover") %>% 
  # summarize and group quadrats by transects
  group_by(year, site, habitat, site_habitat, transect, island_side, substrate) %>% 
  summarise(avg_cov = mean(sum_cover)) -> transect4mean

## Quadrat level island-habitat

```{r make a df for island side/habitat}
macroalgae %>% 
  mutate(macroalgae_sum = rowSums(macroalgae[startMAlg:endMAlg])) %>% 
  select(year, location, site, habitat, site_habitat, transect, quadrat, island_side, macroalgae_sum) %>% 
  full_join(coralAlgDiv) %>% 
  full_join(algae) %>% 
  rowwise() %>% 
  mutate(turf_sum = sum(algal_turf, damselfish_turf),
         CCA_sum = sum(lithophyllum_kotschyanum, crustose_corallines, neogoniolithon_brassica_florida, sporolithon_sp)) %>% 
  select(year, location, site, habitat, site_habitat, transect, quadrat, island_side, turf_sum, CCA_sum, coral, macroalgae_sum) %>% 
  # pivot key columns longer
  pivot_longer(cols = c(turf_sum, CCA_sum, coral, macroalgae_sum),
               names_to = "substrate",
               values_to = "sum_cover") %>% 
  # summarize and group quadrats by island/habitat
  group_by(year, habitat, island_side, substrate) %>% 
  summarise(mean_cov = mean(sum_cover),
            sd_cov = sd(sum_cover, na.rm = TRUE),
            n = n(),
            CI95 = 1.96*(sd_cov)/sqrt(n)) %>% 
  # relevel for graphing
  mutate(island_side = factor(island_side, levels = c("north", "east", "west")),
         habitat = factor(habitat, levels = c("fringing", "backreef", "outer_10", "outer_17"))) -> quadHabIS

```{r plot cover through time}
quadHabIS %>% 
  ggplot(aes(x = year, y = mean_cov, group = substrate, fill = substrate)) +
  geom_ribbon(aes(ymin = mean_cov -CI95, 
                    ymax = mean_cov + CI95),
              alpha  = 0.7) +
  geom_line() +
  theme_bw() +
  scale_fill_manual(values = c(
    moma.colors("Klein", type = "discrete")[2], # CCA
    moma.colors("Klein", type = "discrete")[1], # coral
    moma.colors("Klein", type = "discrete")[5], # macro
    moma.colors("Klein", type = "discrete")[4] # turf
  )) +
  facet_grid(island_side ~ habitat) +
  ylab("Mean cover (%)") +
  xlab("Year")
```

## Transect-level island/habitat

```{r make a df for island side/habitat}
transect4mean %>% 
  group_by(year, habitat, island_side, substrate) %>% 
  summarise(mean_cov = mean(avg_cov),
            sd_cov = sd(avg_cov, na.rm = TRUE),
            n = n(),
            CI95 = 1.96*(sd_cov)/sqrt(n)) %>% 
  # relevel for graphing
  mutate(island_side = factor(island_side, levels = c("north", "east", "west")),
         habitat = factor(habitat, levels = c("fringing", "backreef", "outer_10", "outer_17"))) -> transectHabIS

```{r plot cover through time}
# this color palette:
#moma.colors("Abbott", type = "discrete")

transectHabIS %>% 
  ggplot(aes(x = year, y = mean_cov, group = substrate, fill = substrate)) +
  geom_ribbon(aes(ymin = mean_cov -CI95, 
                    ymax = mean_cov + CI95),
              alpha  = 0.7) +
  geom_line() +
  theme_bw() +
  scale_fill_manual(values = c(
    moma.colors("Klein", type = "discrete")[2], # CCA
    moma.colors("Klein", type = "discrete")[1], # coral
    moma.colors("Klein", type = "discrete")[5], # macro
    moma.colors("Klein", type = "discrete")[4] # turf
  )) +
  facet_grid(island_side ~ habitat) +
  ylab("Mean cover (%)") +
  xlab("Year")
```

## Transect-level habitat

```{r make a df for island side/habitat}
transect4mean %>% 
  group_by(year, habitat, substrate) %>% 
  summarise(mean_cov = mean(avg_cov),
            sd_cov = sd(avg_cov, na.rm = TRUE),
            n = n(),
            CI95 = 1.96*(sd_cov)/sqrt(n)) %>% 
  # relevel for graphing
  mutate(habitat = factor(habitat, levels = c("fringing", "backreef", "outer_10", "outer_17"))) -> transectHab

```{r plot cover through time}
# this color palette:
#moma.colors("Abbott", type = "discrete")

transectHab %>% 
  ggplot(aes(x = year, y = mean_cov, group = substrate, fill = substrate)) +
  geom_ribbon(aes(ymin = mean_cov -CI95, 
                    ymax = mean_cov + CI95),
              alpha  = 0.7) +
  geom_line() +
  theme_bw() +
  scale_fill_manual(values = c(
    moma.colors("Klein", type = "discrete")[2], # CCA
    moma.colors("Klein", type = "discrete")[1], # coral
    moma.colors("Klein", type = "discrete")[5], # macro
    moma.colors("Klein", type = "discrete")[4] # turf
  )) +
  facet_wrap(~habitat) +
  ylab("Mean cover (%)") +
  xlab("Year")
```

## Transect level island side

```{r make a df for island side}
transect4mean %>% 
  group_by(year, island_side, substrate) %>% 
  summarise(mean_cov = mean(avg_cov),
            sd_cov = sd(avg_cov, na.rm = TRUE),
            n = n(),
            CI95 = 1.96*(sd_cov)/sqrt(n)) %>% 
  # relevel for graphing
  mutate(island_side = factor(island_side, levels = c("north", "east", "west"))) -> transectIS

```{r plot cover through time}

transectIS %>% 
  ggplot(aes(x = year, y = mean_cov, group = substrate, fill = substrate)) +
  geom_ribbon(aes(ymin = mean_cov -CI95, 
                    ymax = mean_cov + CI95),
              alpha  = 0.7) +
  geom_line() +
  theme_bw() +
  scale_fill_manual(values = c(
    moma.colors("Klein", type = "discrete")[2], # CCA
    moma.colors("Klein", type = "discrete")[1], # coral
    moma.colors("Klein", type = "discrete")[5], # macro
    moma.colors("Klein", type = "discrete")[4] # turf
  )) +
  facet_grid(~island_side) +
  ylab("Mean cover (%)") +
  xlab("Year")
```

# Examine most abundant algae

Reformat so that the data are easier to summarize:

```{r}
# get ranges to calculate over
startAlg <- grep("acanthophora_spicifera", colnames(algae))
endAlg <- grep("valonia_ventricosa", colnames(algae))

algae %>% 
  pivot_longer(cols = colnames(algae)[startAlg:endAlg],
               names_to = "taxa",
               values_to = "percent") -> algaeLong
```

Get some summary statistics

## Across all time, by habitat:

```{r}
algaeLong %>% 
  group_by(site, habitat, year, transect, island_side, taxa) %>% # get transect summaries
  summarize(Mean_cov = mean(percent), 
            Num_occur = n(), 
            sd_cov = sd(percent)) -> algTaxTrans

Plot

```{r}
algTaxTrans %>% 
  ggplot(aes(x = reorder(taxa, -Mean_cov), y = Mean_cov, fill = habitat)) +
  geom_boxplot() + 
  scale_fill_viridis_d() +
  facet_wrap(~habitat) +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

With just macroalgae:

```{r}
algTaxTrans %>% 
  filter(! taxa %in% c("algal_turf", "crustose_corallines", "cyanophyta", "damselfish_turf", "lithophyllum_kotschyanum", "neogoniolithon_brassica_florida", "sporolithon_sp" )) %>% # this list taken from the cleaning code
  ggplot(aes(x = reorder(taxa, -Mean_cov), y = Mean_cov, fill = habitat)) +
  geom_boxplot() + 
  scale_fill_viridis_d() +
  facet_wrap(~habitat) +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

Not by habitat

```{r}
algTaxTrans %>% 
  filter(! taxa %in% c("algal_turf", "crustose_corallines", "cyanophyta", "damselfish_turf", "lithophyllum_kotschyanum", "neogoniolithon_brassica_florida", "sporolithon_sp" )) %>% # this list taken from the cleaning code
  ggplot(aes(x = reorder(taxa, -Mean_cov), y = Mean_cov)) +
  geom_boxplot() + 
  scale_fill_viridis_d() +
  #facet_wrap(~habitat) +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

@@

```{r}
startAlg <- grep("acanthophora_spicifera", colnames(algae))
endAlg <- grep("valonia_ventricosa", colnames(algae))

# spp matrix
algae %>% 
  mutate(AlgSum = rowSums(algae[startAlg:endAlg])) %>% 
  filter(AlgSum != 0) %>% 
  select(-year, -location, -island_side, -site, -habitat, -site_habitat, -transect, -quadrat, -AlgSum) -> sppMat0

# metadata matrix
algae %>% 
  mutate(AlgSum = rowSums(algae[startAlg:endAlg])) %>% 
  filter(AlgSum != 0) %>%
  select(year, location, site, habitat, site_habitat, transect, quadrat, island_side) %>% 
  # make a year-habitat column for betadisper
  mutate(hab_year = paste0(habitat, "_", year)) -> sppMeta0

@@

- Talking with Lauren, decided want to be smart about which groupings we do for `betadisper()`. Priorities are:
  1. Year-habitat as the group: include turf/cyano/everything
  2. Island side - year - habitat

```{r}
habBetaDisper <- vegan::betadisper(brayMat, sppMeta0$hab_year)

# plot of dispersion test
plot(habBetaDisper, sub = "") # lol
boxplot(habBetaDisper, main = "Habitat", xlab = "") 
```
